---
title: 'DS5220 Project 1 '
author: "Naman Singhal"
date: "Feb 8 2024"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
```

## Instructions

R has excellent visualization capabilities, especially with the `ggplot2` package.  Please read Chapter 3 of  [*R for Data Science [GW],* Garrett Grolemund, Hadley Wickham](https://r4ds.had.co.nz), and complete the exercises below after you finish each section.  Edit the markdown file which came with this html directly.  Make sure to enter your R code in the chunks following each question to demonstrate your answers.  Follow each code block with a text description of your solution.  Answers without demonstration will be given little credit.  Code with no description (if requested) will be given little credit.


## 3.2.4 Exercises

##### 1. Run ggplot(data = mpg). What do you see?  

```{r 'problem 3.2.4.1', eval=FALSE}
ggplot(data = mpg)
```

By running this we see an empty graph because ggplot2 only creates a graph without any mapping layer with geometric function. To fix this we can instead do something like this

```{r Solution, eval=FALSE}
ggplot(mpg) + geom_point(mapping = aes(y = hwy, x = cyl))
```



##### 2. How many rows are in mpg? How many columns? Demonstrate how you obtained your answers using R.

```{r 'problem 3.2.4.2'}
nrow(mpg)
ncol(mpg)
```

We have `r nrow(mpg)` rows and `r ncol(mpg)` columns in the mpg dataset. nrow() function gives us the number of Rows and ncol() gives us number of columns


##### 3. What does the `drv` variable describe? Read the Help Panel in RStudio  by typing `?mpg` in the Console Panel to find out.  (You will see no output from RMarkdown here.)  Produce a description of `drv` by typing `mpg` below.

```{r 'problem 3.2.4.3'}
?mpg
print("drv:f = front-wheel drive, r = rear wheel drive, 4 = 4wd")
```

It describes the type of drive on the vehicle. Field is populated with wither **f** for front wheel drive, **r** for rear and **4** for all 4 wheel drive.



##### 4. Make a scatterplot of `hwy` vs `cyl` using `geom_point`.

```{r 'problem 3.2.4.4'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

In the above geom_point adds a layer for scatter plot on ggplot. AES is used to map coorinates.


##### 5. What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r 'problem 3.2.4.5'}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))

```

From the data points there is no pattern in data, meaning that **class** is weakly correlated with **drv**.


## 3.3.1 Exercises

##### 1. Fix the code in problem 3.3.1.1, and enter it below

```{r problem 3.3.1.1}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color="blue")
```

In the original code of the question the attribute `color` was defined as the parameter of `aes()` which was mapped as a coordinate (aesthetic) to our data point. This is why we saw legends instead of color blue. To make color blue `color` should be a parameter of `geom_point()`.

##### 2. Which variables in `mpg` are categorical? Which variables are continuous? (Hint: type `?mpg` in the Console Panel to read the documentation for the dataset in the Help Panel). How can you see this information when you run `mpg`?

```{r problem 3.3.1.2}
mpg
```

`manufacturer`, `model`, `year`, `trans`, `drv`, `fl`, and `type` are all the features with categorical data. `displ`, `cyl`, `cty`, and `hwy` are continuous.

##### 3. Using your R code for 3.3.1.1, map a continuous variable to `color`, `size`, and `shape`. How does the aesthetic `shape` behave differently for mappings to `fl` and `displ`?

```{r problem 3.3.1.3a}
# For color
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=displ))

# For Size
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size=displ))

# For Shape
# here we use the shape parameter for geom_point()
# This is not throwing error but it is useless because we don't have legends.
# geom_point here has taken the continous variable and scaled it to integer values of the shapes.
ggplot(data = mpg) + 
   geom_point(mapping = aes(x = displ, y = hwy), shape=mpg$displ, show.legend = TRUE)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape=fl))

# ggplot(data = mpg) + 
#   geom_point(mapping = aes(x = displ, y = hwy, shape=mpg$displ), )

# AES() does not have auto scale. Shapes cannot be continous as shapes dont have magnitude. While Color and Size have a magnitude which is continous.
# Caused by error in `scale_f()`:
# ! A continuous variable cannot be mapped to the shape aesthetic
# ℹ choose a different aesthetic or use `scale_shape_binned()`

```

Mapping `displ` to `color` and `size` produce meaningful results. Mapping `displ` to `shape` produces a warning. Assigning the aesthetic `shape` to `fl` produces the following.

```{r problem 3.3.1.3b}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape=fl), color="blue")

```


##### 4. What happens if you map the same variable to multiple aesthetics? Use an example from your answers to problem 3.3.1.3.
```{r}
ggplot(data = mpg) + geom_point(aes(x = displ, y = hwy, color = cty, size = cty))
# Multi aesthetics adds multiple dimensions to the plot

```


##### 5. What does the `stroke` aesthetic do? What shapes does it work with? (Hint: use`?geom_point`) Try it with `shape=21` and `stroke=displ` in your  code from 3.3.1.1.

The `stroke` parameter is an aesthetic in `geom_point` which sets the thickness of the point's border. `stroke` seems to be working with most of the shapes.

```{r problem 3.3.1.5}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color="blue", shape=21, stroke=mpg$displ)
```


##### 6.What happens if you map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`?  Try this by modifying your code in problem 3.3.1.1.

This is using the result of `displ<5` calculation and using that as aesthetics

```{r problem 3.3.1.6}

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color=displ<5))

```

